Type Evolution in Programming Systems

### 2024-05-25 paper.tex
Finish Elm migration, Entity evolution

\usepackage{csquotes}
\usepackage{changepage}
\usepackage{multicol}
\usepackage{ccicons}
\usepackage[many]{tcolorbox}
\usepackage{tikz-cd}
\usepackage{float}

@@

\keywords{schema change, database migration, live programming, local-first programming}

@@

\begin{abstract}
Schema change is an unsolved problem in both live programming and local-first software. We include
in schema change any change to the expected shape of data, whether that is expressed explicitly
in a database schema or type system, or whether those expectations are implicit in the behavior
of the code.

Schema changes during live programming can create a mismatch between the code and
data in the running environment. Correspondingly, Schema changes in local-first programming can
create mismatches between data in different replicas, and between data in a replica and the code
colocated with it. In all of these situations the problem of schema change is to migrate or
translate existing data in coordination with changes to the code.

This paper contributes a set of concrete scenarios involving schema change that are offered as
challenge problems to the live programming and local-first communities. We hope that these
problems will spur progress by providing concrete objectives and a basis for comparing alternative
solutions.

\noindent\textbf{Context}: \emph{What is the broad context of the work? What is the importance of the general research area?}
Schema change is an unsolved problem in live and local-first programming that is poorly understood and calls for a more fundamental analysis. It is particularly important as more programming
is done in programming systems where programs may be changed by their users and as more programming
is local-first, meaning that such changes need to be synchronized.

\noindent\textbf{Inquiry}: \emph{What problem or question does the paper address? How has this problem or question been addressed?}
Specific systems have specific solutions, but what is lacking is a general framework for thinking
about schema change and also a good suite of challenge problems to motivate and evaluate new solutions.

\noindent\textbf{Approach}: \emph{What was done that unveiled new knowledge?}
We collect challenge problems from a range of different kinds of (primarily) live and local-first
application contexts. We develop a framework for thinking about program change and use it to
present our challenge problems in a systematic way.

\noindent\textbf{Knowledge}: \emph{What new facts were uncovered? What new capabilities are enabled by the work?}
We understand change in programs in a novel way, using a two-dimensional framwork that looks at
the different pace layers in a program and at the different ways in which change is propagated
across program variants.

\noindent\textbf{Grounding}: \emph{What argument, feasibility proof, artifacts, or results and evaluation support this work?}
We use the framework to provide a framing for a range of challenge problems derived from both
existing literature and our own expertise.

\noindent\textbf{Importance}: \emph{Why does this work matter?}
This paper contributes a set of concrete scenarios involving schema change that are offered as
challenge problems to the live programming and local-first communities. We hope that these
problems will spur progress by providing concrete objectives and a basis for comparing alternative
solutions.

@@

@@

@@

\section{Live programming in the Elm architecture}
\label{SECT:elm}

The first challenge problem we recast in terms of our conceptual framework is live programming
of user interfaces based on the Elm architecture. In this model, a reactive web application is
structured in terms of current state and events that affect the state. The implementation then
consists of two functions called \texttt{update} and \texttt{render}:

@@

\noindent
Some programming systems that use the Elm architecture support hot reloading when the
implementation of the \texttt{render} or \texttt{update} function changes, but they typically
discard state and restart the application when types are changed. A more effective programming
system based on live programming would allow updates to both the code of the two functions and
the structure of the two types.

\begin{table}[t]
\begin{tabular}{lll}\toprule
{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
add & case to \texttt{Event} & requires adding corresponding case to \texttt{update} \\
remove & case from \texttt{Event} & remove unused code from \texttt{update} \\
add & field to \texttt{State} & migrate state value and initialize field \\
remove & field from \texttt{State} & migrate state (assuming field unused) \\
modify & structure of \texttt{State} & migrate state value and edit code accordingly \\
\bottomrule
\end{tabular}
\vspace{0.3em}
\caption{Changes to types and how a programming system should handle them}
\label{TBL:elmchanges}
\vspace{-1em}
\end{table}

\challenge{Live Type Evolution of State and Events}
Assume that we have a running TODO list with the above state and events. To test the application,
the programmer has already created a number of items and so there is a value of the
\texttt{State} type that represents the current state of the application such as:

@@

More generally, there is a number of edits to the types that the programmer may want to do
without restarting the application. The edits and the desired handling in the programming
system are summarized in Table~\ref{TBL:elmchanges}. Modifying the \texttt{Event} type does not
require data migration (assuming that the change does not happen while there are unprocessed
events). It may result in unused code or missing cases in \texttt{update} that need to be addressed.
Modifying \texttt{State} ranges from relatively simple problems (e.g., adding a new field with a
default value) to challenging case. In particular, if a programer refactors \texttt{State} to
semantically equivalent type, the programming system should, in principle, be able to automatically
migrate both the original value and implementation of both functions to match the new type.

\textcolor{red}{TODO: The rest needs more editing}

\frameworkbox{
In this
}

\subsection{Goals}
As above, the key requirement from the live programming perspective is to ``do minimal harm''. In this case, this results in the following goals:

\begin{itemize}
\item When migrating the application state, this needs to be done automatically and the system should strive to produce new state that is as near as possible to the previous state.
\item The solution may rely on structure editing so that the system has access to a high-level logical description of the edits performed by the user.
\item The system needs to handle the case when the type structure diverges from the code structure. This can be addressed in various ways (transform code, add error handlers, etc.) but it is desirable to avoid "breaking" the \texttt{render} function as this would make the new application state impossible to see.
\end{itemize}

\subsection{Remarks}
It is worth noting that the particularly difficult aspect of this challenge, i.e., the case where the structure of state is refactored to an equivalent one, is related to the Extract Entity challenge.

\newpage
\textcolor{red}{TODO: The rest needs more editing}

@@

\section{Challenge goals}
\textcolor{red}{Text from the Extract Entity challenge that maybe should be lifted up to the top-level discussion of goals}

The Extract Entity challenge for live programming is to provide interactive operations in the programming environment that will perform the schema change on live code and data. The data can be in any form desired, including relations, objects, or JSON. The code can be a simple CRUD UI that also provides the shipping department its view of unshipped orders. The central problem is to coordinate the schema change with code edits/refactorings to keep the system executing live and correctly. Solutions should try to minimize:

\begin{enumerate}
  \item The number and complexity of commands or UI affordances that must be introduced.
  \item Interruption to live execution of the code.
\end{enumerate}

The Extract Entity challenge for local-first software is similar to the live programming challenge except that the context is an application that offers similar functionality with peer-to-peer collaboration. In this context the schema change does not need to be performed interactively -- a developer can take some time to build, test, and package a solution, perhaps using an API or DSL. The hard part comes when the schema change is to be deployed across all the replicas. This deployment must synchronize code and data upgrades (or decouple them as in Cambria~\cite{Cambria}). The deployment must also deal with migrating ``in flight'' operations so that all replicas converge on the same state without data loss, and ideally without centralized coordination. A radical approach could try to incorporate schema change operations into the underlying datastore/CRDT itself and integrate code as well, but layered solutions are welcome too.

\subsection{Goals}
Solutions should try to minimize:

\begin{enumerate}
  \item The need for users to manually intervene.
  \item Possibilities that developer-written code is subtly incorrect in edge cases, preferably by reducing the need for developer-written code.
  \item The need for a centralized coordinator.
  \item How long replicas might be partitioned until resynchronization is acquired.
\end{enumerate}

@@

\begin{figure}[H]
  \centering
  \begin{tikzcd}[sep = 6em]
    \textsf{query}
    \arrow[d, "\textsf{refactor}_\alpha"{name=L, description}]
    &
    \textsf{schema}
    \arrow[l, "\textsf{constrains}"{above}]
    \arrow[r, "\textsf{constrains}"{above}]
    \arrow[d, red, "\textsf{change}_\alpha"{name=C, description}]
    &
    \textsf{data}
    \arrow[d, "\textsf{migrate}_\alpha"{name=R, description}]
    \\
    \textsf{query}_\alpha
    &
    \textsf{schema}_\alpha
    \arrow[l, "\textsf{constrains}"{above}]
    \arrow[r, "\textsf{constrains}"{above}]
    &
    \textsf{data}_\alpha
    \arrow[Rightarrow, from=C, to=L, "\textsf{propagate}"{above}]
    \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
  \end{tikzcd}
  \caption{Extract Entity}
\end{figure}

\begin{figure}[H]
  \centering
  \begin{tikzcd}[column sep = large, row sep = 5em]
    &
    \textsf{schema}
    \arrow[ld, red, "\textsf{change}_\alpha"{name=L2, near start, description}]
    \arrow[rr, "\textsf{constrains}"]
    \arrow[dd, red, "\textsf{change}_\beta"{name=L, near start, description}]
    &
    &
    \textsf{data}
    \arrow[dd, "\textsf{migrate}_\beta"{name=R, near start, description}]
    \arrow[Rightarrow, from=L, to=R]
    \\
    \textsf{schema}_\alpha
    %\arrow[dd, "\textsf{change}_\beta"{name=L3, near start, description}]
    \arrow[rr, crossing over]
    &
    &
    \textsf{data}_\alpha
    \arrow[from=ur, crossing over, "\textsf{migrate}_\alpha"{name=R2, near start, description}]
    \arrow[Rightarrow, from=L2, to=R2, crossing over, "\textsf{propagate}"{above}]
    &
    \\
    &
    \textsf{schema}_\beta
    \arrow[ld, "\mathit{rebase}\left(\textsf{change}_\alpha {\,,\;} \textsf{change}_\beta\right)"{name=L4, near start, description}]
    \arrow[rr]
    &
    &
    \textsf{data}_\beta
    \arrow[ld, "\textsf{migrate}_{\alpha\beta}"{name=R4, near start, description}]
    \arrow[Rightarrow, from=L4, to=R4]
    \\
    \textsf{schema}_{\alpha\beta}
    \arrow[rr]
    &
    &
    \textsf{data}_{\alpha\beta}
    %\arrow[from=uu, crossing over, "\textsf{migrate}_\beta"{name=R3, near start, description}]
    %\arrow[Rightarrow, from=L3, to=R3, crossing over]
    &
  \end{tikzcd}
  \caption{Mergeing divergent changes}
\end{figure}

\begin{figure}[H]
  \centering
  \begin{tikzcd}[column sep = large, row sep = 5em]
    &
    \textsf{schema}
    \arrow[ld, red, "\textsf{change}_\alpha"{name=L2, near start, description}]
    \arrow[rr, "\textsf{constrains}"]
    \arrow[dd, "\textsf{identity}"{near start, description}]
    &
    &
    \textsf{data}
    \arrow[dd, red, "\textsf{update}_\beta"{near start, description}]
    \\
    \textsf{schema}_\alpha
    \arrow[rr, crossing over]
    &
    &
    \textsf{data}_\alpha
    \arrow[from=ur, crossing over, "\textsf{migrate}_\alpha"{name=R2, near start, description}]
    \arrow[Rightarrow, from=L2, to=R2, crossing over, "\textsf{propagate}"{above}]
    \arrow[rr, red, crossing over, "\textsf{update}_\gamma"{near start, description}]
    &
    &
    \textsf{data}_{\alpha\gamma}
    \arrow[dd, "\mathit{rebase}\left(\textsf{update}_\beta {\,,\;} \textsf{update}_\gamma\right)"{near start, description}]
    \\
    &
    \textsf{schema}
    \arrow[ld, "\textsf{change}_\alpha"{name=L4, near start, description}]
    \arrow[rr]
    &
    &
    \textsf{data}_\beta
    \arrow[ld, "\textsf{migrate}_{\alpha\beta}"{name=R4, near start, description}]
    \arrow[Rightarrow, from=L4, to=R4]
    \\
    \textsf{schema}_\alpha
    \arrow[rr]
    &
    &
    \textsf{data}_{\alpha\beta}
    &
    &
    \textsf{data}_{\alpha\beta\gamma}
  \end{tikzcd}
  \caption{Conference Organizer merging schema and divergent data updates}
\end{figure}

\section{Related Work}

@@

\section{Extract/Absorb/Split/Merge Entity}
\textcolor{red}{Maybe called \textit{Individuation}? Maybe called \textit{Entity Evolution}?}

The Acme Corporation needs to record orders from various customers for various products. The simplest implementation is a spreadsheet with a row for each order and columns for information about the customer and product. It might look like this:

\begin{table}[h!]
  \sffamily
  \small
  % \centering
  \begin{tabular}{ |r|l|r|r|l|l|}
   \hline
   order\_id & item & quantity & ship\_date & customer\_name & customer\_address \\
   \hline \hline
   1 & Anvil & 1 & 2/3/23 & Wile E Coyote & 123 Desert Station \\
   \hline
   2 & Dynamite & 2 & & Daffy Duck & White Rock Lake \\
   \hline
   3 & Bird Seed & 1 & & Wile E Coyote & 123 Desert Station \\
   \hline
  \end{tabular}
  \caption{Orders}
\end{table}

The shipping department filters this table on blank ship dates to see what they need to ship. But after a while the orders department realizes they are wasting effort duplicating the address for new order from an old customer. And when the customer's address changes they have to go back and edit all of their orders. What is needed is two tables, one with orders that links to one with customers, like this:

\begin{table}[h!]
  \sffamily
  \small
  % \centering
    \begin{tabular}{ |r|l|l|}
      \hline
      customer\_id & customer\_name & customer\_address \\
      \hline \hline
      1 & Wile E Coyote & 123 Desert Station \\
      \hline
      2 & Daffy Duck & White Rock Lake \\
      \hline
    \end{tabular}
  \caption{Customers}
\end{table}

\begin{table}[h!]
  \sffamily
  \small
  % \centering
    \begin{tabular}{ |r|l|r|r|r|}
     \hline
     order\_id & item & quantity & ship\_date & customer\_id \\
     \hline \hline
     1 & Anvil & 1 & 2/3/23 & 1 \\
     \hline
     2 & Dynamite & 2 & & 2  \\
     \hline
     3 & Bird Seed & 1 & & 1 \\
     \hline
    \end{tabular}
  \caption{Orders linking to Customers}
\end{table}

We often encounter this situation when we realize that the attributes of one type of entity should actually belong to a distinct type of entity that will be associated with the first one. Often the motivation is to centralize changes to those attributes in one place. That is what it means to be an \textit{entity}: a referenceable holder of mutable attributes. We call this operation \textit{Extract Entity}.

Unique identifiers (here consecutive numbers) have been assigned to the customers and are referenced from the orders. This allows, for example, a spelling error in a customer address to be fixed in one place. It also allow a customer name or address to change without breaking the connection from all associated orders. Unique identifiers are the essence of entities, whether they are uniquely generated primary keys in a database or an abstraction of a memory address in a programming language.

It is natural to expect every schema evolution to have an inverse, for the change in requirements that triggered an evolution might change back again, and we ought to pay off the technical debt of obsolete design decisions. Research on \textit{Divergence Control}~\cite{Foster2007, herrmann17, chillon22} further suggests operations should be invertible.
The inverse of \textit{Extract Entity} is \textit{Absorb Entity}. It raises the question of how to handle \textit{one-to-many} relationships. For example should absorbing orders into customers do a relational join, returning us back to the starting point, or should it nest orders within customers, as a NoSQL database might do?

The inverse of \textit{Merge Entities} is \textit{Split Entity}, but that can be awkward. Imagine that we didn't have addresses for customers to disambiguate them with and only knew their non-unique names. Then \textit{Extract Entity} will conflate them and discard the information needed to properly invert the operation. Should this information be retained? One could object that we shouldn't have taken orders for people we can't uniquely identify yet. And even if that was necessary it means the customers really aren't distinct entities and shouldn't have been extracted. These are arguments that we shouldn't need \textit{Split Entity}, contradicting the principle that evolution operations should be invertible. It is interesting to observe that evolution in nature can also fail to invert, leaving behind vestigial features. We consider this to be an unsettled issue.

\subsection{Code Coevolution}
A simple example of how code should evolve along with the schema is a SQL query to report all pending orders in the original schema:

The \textit{Extract Entity} operation described earlier should rewrite this query into:

Alternatively the \textit{Extract Entity} operation could create a view in which the original query is executed.

\subsection{Goals}
A solution to this challenge should:
\begin{enumerate}
  \item Define the \textit{Extract Entity} operation.
  \item Define the \textit{Absorb Entity} operation and motivate the tradeoff between joining and nesting.
  \item Define a \textit{Merge Entities} operation or if using automatic deduplication then discuss how to handle mistakes.
  \item Define the \textit{Split Entity} operation or justify why it is not needed.
  \item If proposing a user-facing solution explain the UI affordances invoking these operations.
  \item Extend these operations to the Divergence Control scenario.
\end{enumerate}

@@

\begin{table}[t]
  \centering
\begin{tabular}{lll}\toprule
\textsc{Change} & \textsc{What} & \textsc{How}\\\midrule
add & state & simple  \\
remove & state & structurally simple, but heuristic if state is current \\
rename & state & simple\\
add & variable & migrate class and initialize field \\
remove & variable & migrate class (NB: assumes variable is unused)\\
rename & variable & rename in class, preserving value\\
type change & variable & migrate class and convert or reinitialize value \\
add & transition & simple \\
remove & transition & structurally simple, but heuristic for pending events\\
event change & transition & structurally simple, but heuristic for pending events\\
add/remove & statement & hotswap code at quiescent point of interpreter\\
\bottomrule
\end{tabular}
\caption{Possible program changes and how to deal with them at run time}
\label{TBL:changes}
\end{table}

The space of possible (well-formed\footnote{The abstract syntax schema is not expressive enough to define all static invariants of the language, such as: states must by unique by name, references variables in actions must be declared, etc.}) changes to a state machine are summarized in Table~\ref{TBL:changes}.
The first column indicates the change category, the second the affected kind of object, and third a short description of how to deal with the respective change category. Some notes about this table:
\begin{itemize}
\item all rows with ``simple'' in the third column only require a quiescent~\cite{Tranquility} point in the interpreter loop to update the run-time structure (e.g., as shown in Figure~\ref{FIG:doorsRuntime}).
\item removing a state, however, is only \textit{structurally} simple, since removal is easy, but special care is required if the subject of removal is the current state. In this case, some heuristic is needed, such as: reject the edit, point the current state to the initial state, or some other strategy (e.g., the nearest state, previous state etc.)
\item everywhere a row mentions ``migrate class'' in the third column, data migration is needed: the run-time objects conforming to the old class must be transformed to instances of the updated class, similar to how Smalltalk migrates objects using \lstinline{become:}.
\item removing a transition or changing the event of a transition potentially has to deal with pending events (e.g., in an event queue) expecting such transitions, since such events are now potentially stale. Strategies to deal with this situation include: simply dropping the events, or requiring that such edits cannot be patched at run time when there are pending events.
\item the type change edit of variable requires a strategy for the current value: either discard and reinitialize, or perform value conversion. For instance, if the type change is from boolean to integer, then true could become 1, and false could become 0.
\item note finally that rename variable is mentioned explicitly as a change: this makes it possible to preserve the run-time value of the variable.
\end{itemize}

@@

While the above example is arguably simple, the problem becomes much more challenging when the programs themselves define data types, classes, records, structures etc. Since possibly many instances (values, objects) of such data types may exist at run-time, these all have to be migrated in such a way that programmer experience is minimally disrupted, and that the invariants of said data types is maintained.
Another extension, tying in with the data-oriented examples above, involves refactorings of data types in a program. Typically a refactoring should be behavior preserving, but can it also preserve run-time data? The minimal example is the consistent variable rename rename in Table~\ref{TBL:changes}, which should not have any effect on run-time state.  A more complex refactoring is described in Section~\ref{SECT:elm}.

>>>>

\usepackage{csquotes}
\usepackage{changepage}
\usepackage{multicol}
\usepackage{ccicons}
\usepackage[many]{tcolorbox}
\usepackage{tikz-cd}

@@

\keywords{Type Evolution, Schema Evolution, Live Programming, Local-first Programming}

@@

\begin{abstract}

% Context: What is the broad context of the work? What is the importance of the general research area?
Many recent programming improvements have been enabled by shorter feedback loops. To further
advance the state of programming, we need programming systems that enable rapid collaboration
and make programming more live across the entire development stack. To do this, programming
systems need to preserve state during development and interaction with the programmer.

% Inquiry: What problem or question does the paper address? How has this problem or question been addressed?
The key obstacle to live collaborative programming systems is the problem of \emph{type evolution}.
When programmer changes the structure of a type, programming system needs to co-evolve
existing live data and adapt code that implements the program behaviour to match the new structure.
This problem has been studied in isolation as schema evolution in databases, hot code reloading
in live programming and dynamic software updating, but we lack a unified comprehensive perspective.

% Approach: What was done that unveiled new knowledge?
In this paper, we develop a unfied conceptual framework for analysing the problem of type evolution.
We use the framework to study type evolution in five diverse case studies.
%
% Knowledge: What new facts were uncovered? What new capabilities are enabled by the work?
By looking at the case studies through a unified perspective, we hope to inspire new thinking
about the problem of type evolution and also encourage knowledge transfer across sub-fields.

% Grounding: What argument, feasibility proof, artifacts, or results and evaluation support this work?
Our work takes the form of a sequence of challenge problems drawn from real-world programming
scenario in diverse field such as live programming, database programming and model-driven
development.
%
% Importance: Why does this work matter?
Tackling the problem of type evolution in a unified way, enabled by our new conceptual framework,
will make it possible to design and build a future generation of effective, live and collaborative
programming systems.

@@

% ==================================================================================================
% INTRODUCTION
% ==================================================================================================

@@

% ==================================================================================================
% FRAMEWORK
% ==================================================================================================

@@

% ==================================================================================================
% ELM
% ==================================================================================================

\section{Live Programming for the Elm Architecture}
\label{sec:elm}

The first challenge problem that we present in terms of our conceptual framework is live programming
of user interfaces based on the Elm architecture. In this model, a reactive web application is
structured in terms of current state and events that affect the state. The implementation then
consists of two functions called \texttt{update} and \texttt{render}:

@@

\noindent
Some programming systems that use the Elm architecture support hot reloading when the
implementation of the \texttt{render} or \texttt{update} function changes, but they typically
discard state and restart the application when the \texttt{State} type changes. A more effective
programming system based on live programming would allow updates to both the code of the two
functions and the structure of the two types.

\begin{table}[t]
\begin{tabular}{lll}\toprule
{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
Add & case to \texttt{Event} & Requires adding corresponding case to \texttt{update} \\
Remove & case from \texttt{Event} & Remove unused code from \texttt{update} \\
Add & field to \texttt{State} & Migrate state value and initialize field \\
Remove & field from \texttt{State} & Migrate state (assuming field unused) \\
Modify & structure of \texttt{State} & Migrate state value and edit code accordingly \\
\bottomrule
\end{tabular}
\vspace{0.3em}
\caption{Changes to types and how a programming system should handle them}
\label{tbl:elmchanges}
\vspace{-1em}
\end{table}

\challenge{Live State Type Evolution}
Assume that we have a running TODO list with the above state and events. To test the application,
the programmer has already created a number of items and so there is a value of the
\texttt{State} type that represents the current state of the application such as:

@@

More generally, there is a number of edits to the types that the programmer may want to do
without restarting the application. The edits and the desired handling in the programming
system are summarized in Table~\ref{tbl:elmchanges}. Modifying the \texttt{Event} type does not
require data migration (assuming that the change does not happen when there are unprocessed
events). It may result in unused code or missing cases in \texttt{update} that need to be addressed.
Modifying \texttt{State} ranges from relatively simple problems (e.g., adding a new field with a
default value) to challenging case. In particular, if the programer refactors \texttt{State} to a
semantically equivalent type, the programming system should, in principle, be able to automatically
migrate both the original value and implementation of both functions to match the new type.

\frameworkbox{
The evolving \emph{type} in this case is the \texttt{State} type, \emph{code} is the
program logic implemented in \texttt{update} and \texttt{render} and \emph{data} is the current
state. To support live programming, a local type change requires synchronizing local data to
match the new type, and also a change in code so that it continues working against the new type
structure. In a non-local scenario, programmers would likely want to synchronize types and code,
but not necessarily data. A collaborative programming system would need to support decentralized
and possibly divergent type evolution, but without data synchronization.
}

\subsection{Remarks}
The key requirement from the live programming perspective is to ``do minimal harm''.
The migration of \emph{data} needs to be done automatically and the system should strive
to produce a new state that is as near as possible to the previous state. Migrating
\emph{code} can be done in various ways (transform code, trigger error condition, etc.),
but it is desirable to avoid breaking the \texttt{render} function as this would make the
new application state impossible to visualize.

The specific case where the type is refactored to a semantically equivalent type is related to
the Extract Entity challenge discussed below. Finally, a programming system tackling this
challenge may rely on structure editing so that the system has access to a high-level logical
description of the edits performed by the user.

% ==================================================================================================
% DATABASES
% ==================================================================================================

\section{Entity Evolution in Data-Centric Systems}
Many problems studied in the context of database systems equally apply to programming systems
if we think of them as stateful environments where programs co-exist with their persistent data.
Although schema evolution is well-studied in databses (Section~\ref{sec:related}), past work
does not resolve all problems in our second challenge.

As an example, consider the Acme Corporation that records orders from various customers for
various products. They use a spreadsheet (Figure~\ref{fig:db-orders}) with a row for each order
and columns for information about the customer and product. The shipping department filters this
table on blank ship dates to see what they need to ship. But after a while the orders department
realizes they are wasting effort duplicating the address for new order from an old customer. And
when the customer's address changes they have to go back and edit all of their orders.

\begin{figure}
\begin{adjustwidth}{-2em}{-3em}
\begin{subfigure}[b]{35em}\vspace{0pt}
  \sffamily
  \small
  \begin{tabular}{ |r|l|r|r|l|l|}
     \hline
     oid & item & quantity & ship\_date & customer\_name & customer\_address \\
     \hline \hline
     1 & Anvil & 1 & 2/3/23 & Wile E Coyote & 123 Desert Station \\
     \hline
     2 & Dynamite & 2 & & Daffy Duck & White Rock Lake \\
     \hline
     3 & Bird Seed & 1 & & Wile E Coyote & 123 Desert Station \\
     \hline
  \end{tabular}
  \caption{Spreadsheet recording orders with all associated information}
  \label{fig:db-orders}
\end{subfigure}
\hfill
\\[-0.5em]~\\
\begin{subfigure}[b]{20em}\vspace{0pt}
  \sffamily
  \small
  \begin{tabular}{ |r|l|l|}
    \hline
    cid & customer\_name & customer\_address \\
    \hline \hline
    1 & Wile E Coyote & 123 Desert Station \\
    \hline
    2 & Daffy Duck & White Rock Lake \\
    \hline
  \end{tabular}
  \caption{Customers table after type evolution}
  \label{fig:db-customers}
\end{subfigure}
\hfill
\begin{subfigure}[b]{20em}\vspace{0pt}
  \sffamily
  \small
  \begin{tabular}{ |r|l|r|r|r|}
     \hline
     oid & item & quantity & ship\_date & cid \\
     \hline \hline
     1 & Anvil & 1 & 2/3/23 & 1 \\
     \hline
     2 & Dynamite & 2 & & 2  \\
     \hline
     3 & Bird Seed & 1 & & 1 \\
     \hline
    \end{tabular}
  \caption{Orders linking to Customers, after type evolution}
 \label{fig:db-orders-link}
\end{subfigure}
\end{adjustwidth}
\vspace{0.25em}
\caption{Types and data before and after performing Extract Entity}
\label{fig:db}
\end{figure}

\challenge{Extract Entity}
Acme needs to evolve the above type and data into two tables (Figure~\ref{fig:db-customers}, \ref{fig:db-orders-link}),
one with orders that links to one with customers.  We refer to this as \emph{Extract Entity}
operation, because it introduces a new \emph{entity}, a referenceable holder of mutable attributes.
The operation is needed when we realize that some attributes of one type of entity should belong to
a  distinct type of entity that will be associated with the first one.

The system needs to provide a mechanism for referencing entities, such as by using unique
identifiers (here consecutive numbers). This allows, for example, a spelling error in a customer
address to be fixed in one place. It also allow a customer name or address to change without
breaking the connection from all associated orders. Unique identifiers are the essence of
entities, whether they are uniquely generated primary keys in a database or an abstraction
of a memory address in a programming language.

More generally, entity evolution can happen through operations such as those listed in
Table~\ref{tbl:dbchanges}. Every schema operation should have an inverse as the
change in requirements that triggered an evolution might change back again. Inverse
is also required in research on \emph{divergence control}~\cite{Foster2007, herrmann17, chillon22}.
The inverse of \textit{Extract Entity} is \textit{Absorb Entity}. It raises the question of how
to handle \textit{one-to-many} relationships. For example should absorbing orders into customers
do a relational join, returning us back to the starting point, or should it nest orders within
customers, as a NoSQL database might do?

The inverse of \textit{Merge Entities} is \textit{Split Entity}, but that can be awkward.
Imagine that we didn't have addresses for customers to disambiguate them with and only knew
their non-unique names. Then \textit{Extract Entity} will conflate them and discard the
information needed to properly invert the operation. Should this information be retained?
One could object that we shouldn't have taken orders for people we can't uniquely identify
yet. And even if that was necessary it means the customers really aren't distinct entities
and shouldn't have been extracted. These are arguments that we shouldn't need \textit{Split
Entity}, contradicting the principle that evolution operations should be invertible.
It is interesting to observe that evolution in nature can also fail to invert, leaving
behind vestigial features. We consider this to be an open issue.

\begin{table}[t]
\begin{tabular}{lll}\toprule
{\firamedium Change} & {\firamedium What} & {\firamedium How}\\\midrule
Extract & new type of entity & Migrate data to a new table and add links \\
Absorb & a linked type of entity & Migrate data and remove unneeded table \\
Merge & multiple data items & Automatic de-duplication with mistake handling \\
Split & multiple data items & Interactively, following user guidance\\
\bottomrule
\end{tabular}
\vspace{0.3em}
\caption{Changes to types and how a programming system should handle them}
\label{tbl:dbchanges}
\vspace{-1em}
\end{table}

\frameworkbox{
In a programming system, the \emph{types} to which entity evolution applies can be data tables,
but also objects or JSON. When the type evolves, the system needs to update corresponding
live \emph{data} and \emph{code}. In a non-local scenario, entity evolution
typically concerns permanent data. As a result, the system needs to synchronize types,
data and code across all replicas. Doing so in a way that allows divergence is discussed
in the next section.
}

\subsection{Remarks}
The central problem is to coordinate the type evolution with code edits to keep the system
executing live and correctly. Solutions may infer the desired type evolution by analysing
source code edits, but would typically offer UI affordances for the commands. They should
try to minimize the number and complexity of commands that must be introduced and also the
interruption to live execution of the code.

The Extract Entity challenge in the non-local case (for local-first software) is similar to the
local live programming challenge except that the context is an application that offers similar
functionality with peer-to-peer collaboration. In this context the schema change does not need
to be performed interactively -- a developer can take some time to build, test, and package a
solution, perhaps using an API or DSL. The hard part comes when the schema change is to
be deployed across all the replicas. This deployment must synchronize code and data upgrades
(or decouple them as in Cambria~\cite{Cambria}). The deployment must also deal with
migrating ``in flight'' operations so that all replicas converge on the same state
without data loss, and ideally without centralized coordination. A radical approach
could try to incorporate schema change operations into the underlying datastore
itself (possibly a CRDT) and integrate code as well, but layered solutions are welcome too.

\challenge{Code Co-evolution for Extract Entity}

In a programming system, there is typically also some code for programmatic manipulation
of entities. The code can, for example, implement a simple CRUD UI that also provides the
Acme shipping department its view of unshipped orders. The holistic perspective of programming
systems lets us see this code not as separate from the system, but as its integral part.
The code thus also needs to co-evolve when the entity evolves. An example is a SQL
query to report all pending orders in the original schema:

\noindent
After applying the \textit{Extract Entity} operation as described earlier, the code should
be rewritten into the following query that joins the two linked tables:

\noindent
The same code transformation would need to be applied to queries written over other
data representations such as JSON. The system can also use mechanism such as a database
view and execute the original query over the view. When applying code migrations,
the system should minimize the need for users to manaually intervene. It should also
minimize the possibilities that developer-written code is subtly incorrect in edge cases,
preferably by reducing the need for developer-written code.

\newpage
~
\newpage

@@

@@

  \begin{tikzcd}[sep = huge]
    \textsf{code}_1
    \arrow[d, "\textsf{refactor}"{left},""{name=L}]
    &
    \textsf{schema}_1
    \arrow[l, "\textsf{constrains}"{above}]
    \arrow[r, "\textsf{constrains}"]
    \arrow[d, red, "\textsf{change}"{near start, description}, ""{name=CL, left}, ""{name=CR}]
    &
    \textsf{data}_1
    \arrow[d, ""{name=R, left}, "\textsf{migrate}"{right}]
    \\
    \textsf{code}_2
    &
    \textsf{schema}_2
    \arrow[l, "\textsf{constrains}"{below}]
    \arrow[r, "\textsf{constrains}"{below}] &
    \textsf{data}_2
    \arrow[Rightarrow, from=CL, to=L, "\textsf{propagate}"{below}]
    \arrow[Rightarrow, from=CR, to=R, "\textsf{propagate}"{below}]
  \end{tikzcd}

\section{Related Work}
\label{sec:related}

@@

@@

\begin{table}[t]
  \centering
\begin{tabular}{lll}\toprule
\textsc{Change} & \textsc{What} & \textsc{How}\\\midrule
add & state & simple  \\
remove & state & structurally simple, but heuristic if state is current \\
rename & state & simple\\
add & variable & migrate class and initialize field \\
remove & variable & migrate class (NB: assumes variable is unused)\\
rename & variable & rename in class, preserving value\\
type change & variable & migrate class and convert or reinitialize value \\
add & transition & simple \\
remove & transition & structurally simple, but heuristic for pending events\\
event change & transition & structurally simple, but heuristic for pending events\\
add/remove & statement & hotswap code at quiescent point of interpreter\\
\bottomrule
\end{tabular}
\caption{Possible program changes and how to deal with them at run time}
\label{tbl:smchanges}
\end{table}

The space of possible (well-formed\footnote{The abstract syntax schema is not expressive enough to define all static invariants of the language, such as: states must by unique by name, references variables in actions must be declared, etc.}) changes to a state machine are summarized in Table~\ref{tbl:smchanges}.
The first column indicates the change category, the second the affected kind of object, and third a short description of how to deal with the respective change category. Some notes about this table:
\begin{itemize}
\item all rows with ``simple'' in the third column only require a quiescent~\cite{Tranquility} point in the interpreter loop to update the run-time structure (e.g., as shown in Figure~\ref{FIG:doorsRuntime}).
\item removing a state, however, is only \textit{structurally} simple, since removal is easy, but special care is required if the subject of removal is the current state. In this case, some heuristic is needed, such as: reject the edit, point the current state to the initial state, or some other strategy (e.g., the nearest state, previous state etc.)
\item everywhere a row mentions ``migrate class'' in the third column, data migration is needed: the run-time objects conforming to the old class must be transformed to instances of the updated class, similar to how Smalltalk migrates objects using \lstinline{become:}.
\item removing a transition or changing the event of a transition potentially has to deal with pending events (e.g., in an event queue) expecting such transitions, since such events are now potentially stale. Strategies to deal with this situation include: simply dropping the events, or requiring that such edits cannot be patched at run time when there are pending events.
\item the type change edit of variable requires a strategy for the current value: either discard and reinitialize, or perform value conversion. For instance, if the type change is from boolean to integer, then true could become 1, and false could become 0.
\item note finally that rename variable is mentioned explicitly as a change: this makes it possible to preserve the run-time value of the variable.
\end{itemize}

@@

While the above example is arguably simple, the problem becomes much more challenging when the programs themselves define data types, classes, records, structures etc. Since possibly many instances (values, objects) of such data types may exist at run-time, these all have to be migrated in such a way that programmer experience is minimally disrupted, and that the invariants of said data types is maintained.
Another extension, tying in with the data-oriented examples above, involves refactorings of data types in a program. Typically a refactoring should be behavior preserving, but can it also preserve run-time data? The minimal example is the consistent variable rename rename in Table~\ref{tbl:smchanges}, which should not have any effect on run-time state.  A more complex refactoring is described in Section~\ref{sec:elm}.
